\emph{On Logics Extended with Embedding-Closed Quantifiers}

### 2021-07-15 open_questions.tex
Move open questions to a separate file

\artstatus{am} %%% leave this alone!!  That means you, too!!
%%%%%%%%theorems%%%%%
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

\theoremstyle{plain}
\newtheorem{conjecture}{Conjecture}[section]
\newtheorem{theorem}[conjecture]{Theorem}
\newtheorem{lemma}[conjecture]{Lemma}
\newtheorem{proposition}[conjecture]{Proposition}
\newtheorem{corollary}[conjecture]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[conjecture]{Definition}
\newtheorem{remark}[conjecture]{Remark}
\newtheorem{question}[conjecture]{Question}
\newtheorem{example}[conjecture]{Example}
\newtheorem{examples}[conjecture]{Examples}
\newtheorem{construction}[conjecture]{Construction}

\numberwithin{equation}{section}

\startlocaldefs

\DeclareMathOperator{\FO}{FO}
\DeclareMathOperator{\inv}{inv}
\DeclareMathOperator{\qr}{qr}
\DeclareMathOperator{\hlr}{hlr}
\DeclareMathOperator{\lr}{lr}
\DeclareMathOperator{\Cnt}{Cnt}
\DeclareMathOperator{\BIT}{BIT}
\DeclareMathOperator{\MSO}{MSO}
\DeclareMathOperator{\BNDP}{BNDP}
\DeclareMathOperator{\PG}{PG}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\ran}{ran}
\DeclareMathOperator{\Fun}{Fun}
\DeclareMathOperator{\DetTrCl}{DetTrCl}
\DeclareMathOperator{\dtrcl}{dtrcl}
\DeclareMathOperator{\TrCl}{TrCl}
\DeclareMathOperator{\trcl}{trcl}
\DeclareMathOperator{\LFP}{LFP}
\DeclareMathOperator{\lfp}{lfp}
\DeclareMathOperator{\REACH}{REACH}
\DeclareMathOperator{\emb}{emb}
\DeclareMathOperator{\sub}{sub}
\DeclareMathOperator{\Inj}{inj}
\DeclareMathOperator{\bij}{bij}
\DeclareMathOperator{\Th}{Th}
\DeclareMathOperator{\elem}{elem}
\DeclareMathOperator{\cf}{cf}
\DeclareMathOperator{\cl}{cl}
\DeclareMathOperator{\Str}{Str}
\DeclareMathOperator{\Card}{Card}
\DeclareMathOperator{\ar}{ar}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\frvar}{frvar}
\DeclareMathOperator{\EC}{EC}
\DeclareMathOperator{\PC}{PC}
\DeclareMathOperator{\disjoint}{disjoint}

\endlocaldefs

\date{}

@@

\begin{frontmatter}

	\title{\emph{On Logics Extended with Embedding-Closed Quantifiers}}
	\runtitle{Logics with Embedding-Closed Quantifiers}

	\author{\fnms{Jevgeni}
		\snm{Haigora}
		\corref{}
		\ead[label=e1]{jhaigora@gmail.com}
	}
	\address{
		\printead{e1}\\
	}
	\and
	\author{\fnms{Kerkko}
		\snm{Luosto}
		\ead[label=e2]{Kerkko.Luosto@tuni.fi}
	}
	\address{Faculty of Information Technology and Communication Sciences \\
	  Computing Sciences \\
		Tampere University \\
		FI-33014, Tampere\\
		FINLAND\\
		\printead{e2}\\
	}
	\runauthor{J.~Haigora and K.~Luosto}

\begin{abstract}
	We study first-order as well as infinitary logics
	extended with quantifiers closed upwards under embeddings.
	In particular, we show that if a chain of quasi-homogeneous structures is
	sufficiently long then, in that chain, a given formula of such a logic becomes
	eventually equivalent to a quantifier-free formula.
	We use this fact to produce a number of undefinability results for logics with
	embedding-closed quantifiers.
	We also introduce an Ehrenfeucht-Fra\"iss\'e game that characterizes the
	$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$-equivalence between
	structures,
	where $\mathcal{Q}_{\emb}$ is the class of all embedding-closed quantifiers.
	In conclusion, we provide an application of this game illustrating its use.
\end{abstract}

\begin{keyword}[class=AMS]
	\kwd[Primary ]{03C80}
\end{keyword}

\begin{keyword}
	\kwd{generalized quantifiers} \kwd{embeddings}
	\kwd{Ehrenfeucht-Fra\"iss\'e games}
\end{keyword}

\end{frontmatter}

\section{Introduction}

In this paper we focus our attention on a certain class of model-theoretic
logics whose expressive power is greater than that of first-order logic
(denoted here by $\mathcal{L}_{\omega\omega}$), the central logic in
mathematics.
While $\mathcal{L}_{\omega\omega}$ has well-developed model theory due to its
many convenient properties, it has a downside in that its expressive power is
somewhat limited.
Many natural mathematical statements, for example ``there are infinitely many'',
cannot be expressed in $\mathcal{L}_{\omega\omega}$.
This motivates studying alternative logics.

Mostowski was one of the first to suggest in \cite{Mostowski:1957}
the idea of expanding $\mathcal{L}_{\omega\omega}$ with additional quantifiers.
An example of such an extension is the logic
$\mathcal{L}_{\omega\omega}(Q_{\alpha})$ containing formulas of the form
$Q_{\alpha} x \varphi(x)$ which are interpreted so that
$\mathfrak{A} \vDash  Q_{\alpha} x \varphi(x)$ if and only if there are at least
$\aleph_{\alpha}$ elements $a$ with $\mathfrak{A} \vDash \varphi(a)$.
This idea broadened the notion of quantifier giving rise to many interesting
logics defined in a similar way.
The current definition of generalized quantifier is due to Lindstr\"om
\cite{Lindstrom:1966}. We describe it in more detail in Section \ref{prel}.
A lot of introductory information concerning generalized quantifiers can be also
found in the book \cite{Ebbinghaus:1985}.

In short, every generalized quantifier $Q$ corresponds to some property of
structures that we denote by $\mathcal{K}_Q$.
Suppose $\mathcal{L}$ is a logic closed under substitution and $P$ is a property
not expressible in it. By adding quantifier $Q_P$ to $\mathcal{L}$ we get the
smallest extension of $\mathcal{L}$ satisfying certain closure conditions that
can express $P$.
The properties of the new logic $\mathcal{L}(Q_P)$ can differ substantially from
those of $\mathcal{L}$ and may thus become an interesting object of study.
A justification for studying generalized quantifiers comes among other things
from the fact that any reasonable extension $\mathcal{L}$ of, say, the logic
$\mathcal{L}_{\omega\omega}$,
closed under substitutions, is equivalent to the logic
$\mathcal{L}_{\omega\omega}(\mathcal{Q})$ where $\mathcal{Q}$ is a class of
quantifiers corresponding to some new properties definable in $\mathcal{L}$.

In the present work, we will concentrate on extensions of logics
$\mathcal{L}_{\omega\omega}$, $\mathcal{L}_{\infty \omega}$ and
$\mathcal{L}_{\infty \omega}^{\omega}$ (the finite variable logic) with
generalized quantifiers $Q$ that satisfy the following restriction: for all
structures $\mathfrak{A} \in \Str[\tau_Q]$, if $\mathfrak{A} \in \mathcal{K}_Q$
and $\mathfrak{A}$ is embeddable into $\mathfrak{B}$, then
$\mathfrak{B} \in \mathcal{K}_Q$.
We call such quantifiers \emph{embedding-closed} and denote the class of all
embedding-closed quantifiers by $\mathcal{Q}_{\emb}$.
This is an interesting class of quantifiers to study since many well-known
quantifiers and properties, like cardinality quantifiers $Q_{\alpha}$ for
instance, are closed under embeddings either upwards or downwards which is
essentially the same for our purposes. We present more examples of such
properties in Section \ref{emb-cl_quant}.

Call a structure $\mathfrak{A}$ \emph{quasi-homogeneous} if every isomorphism
between fi\-ni\-te\-ly generated substructures of $\mathfrak{A}$ can be extended
to an embedding of $\mathfrak{A}$ into itself.
This weakens the usual notion of homogeneity which deals with automorphisms
instead of embeddings.
The notion of embedding-closed quantifier arises naturally when we observe in
Theorem \ref{homog} that in order to guarantee that a quasi-homogeneous
structure has quantifier elimination for logic $\mathcal{L}_{\infty\omega}$
extended with a set of quantifiers $\mathcal{Q}$, the quantifiers in
$\mathcal{Q}$ must be closed under embeddings.

In \cite{Dawar:2010}, Dawar and Gr\"adel showed that
$\mathcal{L}_{\infty \omega}^{\omega}$ augmented with finitely many
embedding-closed quantifiers of finite width has a $0$-$1$ law meaning that on
finite structures such a logic can only express properties that hold in almost
all finite structures.
Our aim is to study further limits of the expressive power of logics with
embedding-closed quantifiers that are not implied by a $0$-$1$ law.
These include for example undefinability of properties of infinite structures
and structures with function symbols.
In this article we provide two methods that make it possible. The first method
involves construction of a certain chain of quasi-homogeneous structures.
The second method is based
on the Ehrenfeucht-Fra\"iss\'e game that we develop in order to characterize
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$-equivalence between
structures.
In the article we apply these methods to produce a number of undefinability
results.

The article is structured as follows.
In Section \ref{prel} we introduce preliminary notions.
In Section \ref{emb-cl_quant} we describe basic properties of embedding-closed
quantifiers that will be needed later, and give some examples.
Before moving to our own major results, we show that
$\mathcal{L}^{\omega}_{\infty \omega}(\mathcal{Q})$, where $\mathcal{Q}$ is a
finite set of embedding-closed quantifiers of finite width, has a $0$-$1$ law.
We do this in Section \ref{law}.
The proof concerning the $0$-$1$ law was originally given in \cite{Dawar:2010}.
In Section \ref{chain_section} we introduce the notion of quasi-homogeneity and
show that if a chain of quasi-homogeneous structures is sufficiently long, then
the truth value of a given sentence of a logic with embedding-closed quantifiers
is eventually preserved.
This in turn allows us to obtain some undefinability results.
The section has two subsections, one of which is devoted to the undefinability
of properties of finite structures and another deals with infinite structures.
In Section \ref{game_section} we describe the \emph{embedding game} that
characterizes $\mathcal{L}_{\infty\omega}(\mathcal{Q}_{\emb})$-equivalence of a
given pair of structures.
We close the section with an application of the game that allows us to show that
for each $n<\omega$ there is a first-order sentence of quantifier rank $n$ that
is not expressible by any sentence of
$\mathcal{L}_{\infty\omega}(\mathcal{Q}_{\emb})$ of quantifier rank less than
$n$.

\section{Preliminaries}\label{prel}
The notation for the basic concepts of abstract logics that we use in this paper
is taken from the book \cite{Ebbinghaus:1985}, and we assume that the reader is
familiar with them.
A \emph{vocabulary} $\tau$ consists of \emph{relation}, \emph{function} and
\emph{constant symbols},
\[
	\tau = \{R,\dots,f,\dots,c,\dots\}.
\]
We denote by $\ar(R)$ and $\ar(f)$ the \emph{arities} of relation and function
symbols, respectively.
A \emph{$\tau$-structure} $\mathfrak{A}$ is a sequence
\[
	\mathfrak{A} = (A,R^{\mathfrak{A}},\dots,f^{\mathfrak{A}},\dots,
	c^{\mathfrak{A}},\dots),
\]
where $A$ is a set that we call the \emph{universe} of $\mathfrak{A}$,
and $R^{\mathfrak{A}}\subseteq A^{\ar(R)},\dots$ are interpretations of symbols
of $\tau$.
We denote the class of all $\tau$-structures by $\Str[\tau]$.
We denote the number of free variables of $\varphi$ by $\frvar(\varphi)$.
If $\mathfrak{A}\in\Str[\tau]$ and $\varphi \in \mathcal{L}[\tau]$, then we
write
\[
	\varphi^{\mathfrak{A}} =
	\{\overline{a} \in A^{\frvar(\varphi)} \colon
	\mathfrak{A},\overline{a} \vDash \varphi  \}.
\]

Let $\tau'$ be an extension of $\tau$ and $\mathfrak{A}'$ a $\tau'$-structure.
We say that $\mathfrak{A}$ is the \emph{$\tau$-reduct} of $\mathfrak{A}'$,
and write $\mathfrak{A} = \mathfrak{A}' \upharpoonright \tau$, if $A = A'$ and
$S^{\mathfrak{A}} = S^{\mathfrak{A}'}$ for all $S \in \tau$.

A \emph{literal} is an atomic formula or a negation of an atomic formula.
An \emph{atomic $n$-type} of $\tau$ is a set $\Phi$ of literals of $\tau$ in
variables $x_1,\dots,x_n$ such that there is a $\tau$-structure $\mathfrak{A}$
and $n$-tuple $\overline{a}$ of elements in $A$ with
\[
	\Phi = \{\varphi \colon \mathfrak{A},\overline{a} \vDash \varphi
	\text{ and } \varphi \text{ is a literal of }\tau \}.
\]
If $\mathcal{L}$ is a logic, $\Phi$ an atomic type and there is an
$\mathcal{L}$-formula $t$ logically equivalent to $\bigwedge \Phi$,
then $t$ is also called an atomic type.

\begin{lemma}\label{atomic_types}
Let $\vartheta$ be a quantifier-free $\tau$-formula in $n$ free variables.
Then there exists a set $T$ of atomic $n$-types of $\tau$ such that
\[
	\vDash \vartheta \leftrightarrow
	\bigvee_{\Phi\in T} \bigwedge_{\varphi\in\Phi} \varphi.
\]
\end{lemma}

In this article we will consider the following logics.
We assume that the reader is familiar with the \emph{first-order logic}
$\mathcal{L}_{\omega\omega}$ and the related notions.
Let $\kappa$ be an infinite cardinal.
The logic $\mathcal{L}_{\kappa\omega}$ is allowed to have conjunctions and
disjunctions over sets of formulas of cardinality less than $\kappa$.
The logic $\mathcal{L}_{\infty\omega}$ can have conjunctions and disjunctions
over arbitrary sets of formulas.
Formulas of the logic $\mathcal{L}_{\infty\omega}^{\omega}$
(\emph{finite variable logic}) are exactly those of
$\mathcal{L}_{\infty \omega}$ that use at most finite number of variables.

Suppose $\mathcal{L}$ is a logic and $\tau,\sigma$ are vocabularies where
$\sigma$ has only relation symbols.
An \emph{$\mathcal{L}$-interpretation} of $\sigma$ in $\tau$ is a sequence
$(\Psi,(\psi_R)_{R\in\sigma})$, where each $\psi_R$ is an
$\mathcal{L}[\tau]$-formula that has exactly $\ar(R)$ free variables,
and $\Psi$ is a function $\Str[\tau] \rightarrow \Str[\sigma]$ such that
for each $\mathfrak{A}\in\Str[\tau]$, the universe of $\Psi(\mathfrak{A})$ is
$A$ and
\begin{equation*}
	R^{\Psi(\mathfrak{A})} = \{\overline{a} \in A^{\ar(R)} \colon
	\mathfrak{A}\vDash\psi_R(\overline{a}) \}
\end{equation*}
for all $R\in\sigma$.

Let $C \subseteq \Str[\sigma]$ be a class of structures closed under
isomorphism.
The logic $\mathcal{L}_{\kappa\omega}(Q_C)$ is the smallest extension of
$\mathcal{L}_{\kappa\omega}$ closed under negation, conjunctions and
disjunctions of cardinality $<\kappa$ and application of the existential
quantifier $\exists$ such that  for all
$\mathcal{L}_{\kappa\omega}(Q_C)$-interpretations
$(\Psi,(\psi_R)_{R\in \sigma})$ there is a
$\mathcal{L}_{\kappa\omega}(Q_C)[\tau]$-formula $\chi$ such that
\begin{equation}\label{gen_quant_formula}
	\mathfrak{A} \vDash \chi \Leftrightarrow \Psi(\mathfrak{A})\in C
\end{equation}
for all $\mathfrak{A} \in \Str[\tau]$.
In a similar way we define the logics $\mathcal{L}_{\infty \omega}(Q_C)$
and $\mathcal{L}^{\omega}_{\infty\omega}(Q_C)$.
We say that $Q_C$ is the \emph{generalized quantifier} corresponding to the
class $C$, and write $Q_C(\overline{x}_R\psi_R)_{R\in\sigma}$ for the formula
$\chi$ in \eqref{gen_quant_formula}.
The class $C$ is called the \emph{defining class} of $Q_C$, and $\sigma$ the
\emph{vocabulary} of $Q$.
Sometimes we denote a quantifier just by the symbol $Q$, its defining class by
$K_Q$, and the vocabulary of $Q$ by $\tau_Q$.
We say that $|\tau_Q|$ is the \emph{width} of $Q$.
In a similar way, we can also define the extension of a given logic
$\mathcal{L}$ with a \emph{class} $\mathcal{Q}$ of quantifiers, instead of just
one, which we denote by $\mathcal{L}(\mathcal{Q})$.

\begin{lemma}\label{interpr}
Let $\mathcal{Q}$ be a possibly empty class of quantifiers, $\kappa$ an infinite
cardinal and $\mathcal{L} = \mathcal{L}_{\kappa \omega}(\mathcal{Q})$ or
$\mathcal{L} = \mathcal{L}^{\omega}_{\infty\omega}(\mathcal{Q})$.
Suppose $(\Psi,(\psi_R)_{R\in \sigma})$ is an $\mathcal{L}$-interpretation of
$\sigma$ in $\tau$.
Then for each $\mathcal{L}[\sigma]$-formula $\varphi$ there is an
$\mathcal{L}[\tau]$-formula $\varphi^*$ such that
\[
	\mathfrak{A},\overline{a} \vDash \varphi*
	\Leftrightarrow \Psi(\mathfrak{A}),\overline{a} \vDash \varphi
\]
for all $\mathfrak{A} \in \Str[\tau]$ and tuples $\overline{a}$ of elements in
$A$.
\end{lemma}
\begin{proof}
Replace all atomic subformulas $R(\overline{x})$ of $\varphi$ with formulas
$\psi_R(\overline{x})$ to get $\varphi^*$.
\end{proof}

\begin{lemma}\label{prel_lemma}
Let $\mathcal{Q}_0$ be a possibly empty class of quantifiers, $\kappa$ an
infinite cardinal and $\mathcal{L} = \mathcal{L}_{\kappa \omega}(\mathcal{Q}_0)$
or $\mathcal{L} = \mathcal{L}^{\omega}_{\infty\omega}(\mathcal{Q}_0)$.
Suppose $\mathcal{Q}_1$ is a class of quantifiers such that for every
$Q\in\mathcal{Q}_1$ its defining class $\mathcal{K}_Q$ is definable in
$\mathcal{L}$.
Then $\mathcal{L}(\mathcal{Q}_1) \equiv \mathcal{L}$.
\end{lemma}
\begin{proof}
Let $\tau$ be a vocabulary and consider a formula
$Q(\overline{x}_R\varphi_R)_{R\in\tau_Q}$ with
$Q \in \mathcal{Q}_1$ and $\varphi_R \in \mathcal{L}[\tau]$ for all $R\in\tau_Q$.
Let $\psi \in \mathcal{L}[\tau_Q]$ be the sentence defining $\mathcal{K}_Q$.
Then
\[
	\vDash Q(x_R\varphi_R)_{R\in\tau_Q} \leftrightarrow \chi
\]
where $\chi$ is the formula got from $\psi$ by substituting every atomic
subformula $R(\overline{x})$ of $\psi$ with $\varphi_R(\overline{x})$.
\end{proof}

\section{Embedding-Closed Quantifiers}\label{emb-cl_quant}

\begin{definition}
Let $\mathfrak{A}$ and $\mathfrak{B}$ be structures of the same vocabulary $\tau$.
An injection $f \colon A \rightarrow B$ is an \emph{embedding of $\mathfrak{A}$
into $\mathfrak{B}$} if
\begin{enumerate}
\item $f(c^{\mathfrak{A}}) = c^{\mathfrak{B}}$ for all constant symbols
$c \in \tau$,
\item $\overline{a} \in R^{\mathfrak{A}} \Leftrightarrow f\overline{a} \in R^{\mathfrak{B}}$
for all relation symbols $R \in \tau$ and tuples $\overline{a}$ in $A$,
\item $fF^{\mathfrak{A}}(\overline{a}) = F^{\mathfrak{B}}(f\overline{a})$ for
all function symbols $F \in \tau$ and tuples $\overline{a}$ in $A$.
\end{enumerate}
Here and later in this article we denote by $f$ also the natural extensions of
$f$ to functions $A^n \rightarrow B^n$.

The notation $\mathfrak{A} \leq \mathfrak{B}$ means that $\mathfrak{A}$ is
embeddable into $\mathfrak{B}$, and $\mathfrak{A} < \mathfrak{B}$ means that
$\mathfrak{A} \leq \mathfrak{B}$ but $\mathfrak{B} \nleq \mathfrak{A}$.
We say that a sequence $(\mathfrak{A}_{\alpha})_{\alpha < \gamma}$ of
$\tau$-structures is a \emph{chain} if
$\mathfrak{A}_{\alpha} \leq \mathfrak{A}_{\beta}$ whenever $\alpha \leq \beta$.
We say that a class $C$ of $\tau$-structures is an \emph{antichain} if
$\mathfrak{A} \leq \mathfrak{B}$ is never true for any distinct structures
$\mathfrak{A}, \mathfrak{B} \in C$.

A class $K$ of $\tau$-structures is \emph{embedding-closed} if
$\mathfrak{A} \in K$ and $\mathfrak{A} \leq \mathfrak{B}$ imply
$\mathfrak{B} \in K$. We say that a quantifier $Q$ is embedding-closed if its
defining class is embedding-closed.
We denote by $\mathcal{Q}_{\emb}$ the class of all embedding-closed quantifiers.
\end{definition}

\begin{lemma}\label{preservation}
Let $\tau$ be a vocabulary,
$(\varphi_{\alpha} )_{\alpha < \kappa}$ quantifier-free $\tau$-formulas and $Q$
an embedding-closed quantifier of width $\kappa$.
The formula $Q(\overline{x}_{\alpha} \varphi_{\alpha})_{\alpha < \kappa}$ is
preserved by embeddings.
\end{lemma}
\begin{proof}
Let $\mathfrak{A}$ and $\mathfrak{B}$ be $\tau$-structures.
Suppose that
$(\mathfrak{A},\overline{a}) \vDash Q(\overline{x}_{\alpha} \varphi_{\alpha})_{\alpha < \kappa}$
and $f \colon A \rightarrow B$ is an embedding.
Then $f$ is also an embedding of
$(A,(\varphi_{\alpha}^{\mathfrak{A},\overline{a}})_{\alpha < \kappa})$ into
$(B,(\varphi_{\alpha}^{\mathfrak{B},f\overline{a}})_{\alpha < \kappa})$ since
quantifier-free formulas are preserved by embeddings, so
$(\mathfrak{B},f\overline{a}) \vDash Q(\overline{x}_{\alpha} \varphi_{\alpha})_{\alpha < \kappa}$
since $Q$ is embedding-closed.
\end{proof}

Note that instead of requiring the quantifiers to be closed upwards under
embeddings, we could use the downwards closure to get an equivalent class of
quantifiers.
Call a quantifier $Q$ \emph{substructure-closed} if from
$\mathfrak{A} \in K_Q$ and $\mathfrak{B} \leq \mathfrak{A}$ follows
$\mathfrak{B} \in K_Q$, and denote the class of all substructure-closed
quantifiers by $\mathcal{Q}_{\sub}$.
The expressive power of $\mathcal{Q}_{\sub}$ is clearly the same as that of
$\mathcal{Q}_{\emb}$ since the complement $Q^*$ of an embedding-closed quantifier
$Q$ is substructure-closed and vice versa, so
\[
	\mathfrak{A} \vDash Q(\overline{x}_{\alpha}\varphi_{\alpha})_{\alpha < \kappa}
	\Leftrightarrow  \mathfrak{A} \vDash
	\neg Q^*(\overline{x}_{\alpha}\varphi_{\alpha})_{\alpha < \kappa}.
\]

Next we present some examples of well-known properties and quantifiers that are
either embedding-closed or are definable in the logic
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$.
We use notation $Q^{\cl}$ to denote the closure of the quantifier $Q$ under
embeddings.
In other words $Q^{\cl}$ is the smallest embedding-closed quantifier containing $Q$.

\begin{examples}
\begin{enumerate}
\item Let $\tau = \{U\}$ be a vocabulary consisting of a single unary relation
symbol.
The existential quantifier $\exists$ corresponds to the class of structures
$\{\mathfrak{A} \in \Str[\tau] \colon U^{\mathfrak{A}} \ne \emptyset \}$,
so it is embedding-closed.

\item Let $\tau$ be the same as above and $\alpha$ an ordinal. The defining
class of the \emph{cardinality quantifier} $Q_{\alpha}$ is
$\{\mathfrak{A} \in \Str[\tau] \colon |U^{\mathfrak{A}}| \geq \aleph_{\alpha} \}$
which is clearly embedding-closed.

\item For each $n < \omega$, let $\sigma_n = \{M_n\}$ be a vocabulary consisting
of a single $n$-ary relation symbol.
The \emph{Magidor-Malitz} quantifier $Q_{\alpha}^n$, whose defining class is
\[
	\{ \mathfrak{A} \in \Str[\sigma_n] \colon
	\text{ there is } C \subseteq A \text{ with } |C| \geq \aleph_{\alpha}
	\text{ and } C^n \subseteq M_n^{\mathfrak{A}}\},
\]
is embedding-closed.

\item The well-ordering quantifier $Q^W$, whose defining class is the class of
all well-orders, is substructure-closed, so by our earlier remark it can be
defined by embedding-closed quantifiers.

\item The equivalence quantifier $Q^E_{\alpha}$, whose defining class consists
of all structures $(A,E)$ where $E$ is an equivalence relation on $A$ with at
least $\aleph_{\alpha}$ equivalence classes, is not embedding-closed.
Nonetheless, it can be defined by the sentence
\[
	(Q^E_{\alpha})^{\cl}xyE(x,y) \land \text{``}E \text{ is an equivalence relation''}
\]
in $\mathcal{L}_{\omega\omega}(Q)$ where $Q = (Q^E_{\alpha})^{\cl}$ is
embedding-closed.

\item Many graph properties are embedding- or substructure-closed. Examples
include $k$-colourability, being a forest, completeness, planarity, having a
cycle, and many others.

\item
This is an example of a graph property that is not embedding- or
substructure-closed but is however definable in $\mathcal{L}_{\omega\omega}(Q)$
for an em\-bed\-ding-closed quantifier $Q$.
The property in question is connectedness of a graph.
Let $\sigma = \{R,B,E\}$ be the vocabulary of partially
coloured graphs where symbols $R$ and $B$ stand for colours \emph{red} and
\emph{blue}.
Let $C \subseteq \Str[\sigma]$ consist of all the partially blue-red-coloured
graphs $G$ in which for every blue-red pair $(x,y)$ of vertices there is a path
between $x$ and $y$ and $R^G$, $B^G$ are not empty.
Put $D = C^{\cl}$.
Then for all graphs $G$,
\[
	G \vDash \forall xy Q_Dstuv(s=x,t=y,E(u,v))
\]
if and only if $G$ is connected.
\end{enumerate}
\end{examples}

As we will show below, there exist properties not definable in
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$.
These include among others equicardinality of sets (Example \ref{Hartig}), and
completeness and cofinality of an ordering (Examples \ref{completeness} and
\ref{cofinality}).

\begin{remark}
In addition to being closed under embeddings, we can imagine vari\-ous other
interesting closure conditions for quantifiers, for example, being closed under
homomorphisms.
The rest of this article will be restricted to the study of
embedding-closed quantifiers only.
However, we would like to mention here that it can be shown
that any embedding-closed property can
be defined in a logic with only homomorphism-closed quantifiers.
Thus, requiring quantifiers to be closed under homomorphisms is not essentially
more restrictive than the requirement that they are closed under embeddings.
\end{remark}

\section{$0$-$1$ Law}\label{law}

\begin{definition}
Let $\tau$ be a finite relational vocabulary, and suppose
$P \subseteq \Str[\tau]$ is a property of finite $\tau$-structures.
We say that the \emph{asymptotic probability of $P$} is $r$, and write
$\mu(P) = r$, if
\[
	\lim_{n\rightarrow\infty} \frac{|P \cap \Str_n[\tau]|}{|\Str_n[\tau]|} = r
\]
where $\Str_n[\tau]$ is the set of all $\tau$-structures with domain
$\{0,\dots,n-1\}$, and $|\cdot|$ tells cardinality of a set.

We say that a logic $\mathcal{L}$ \emph{has a $0$-$1$ law} if for every property
$P$ of finite relational structures expressible in $\mathcal{L}$ we have either
$\mu(P) = 1$ or $\mu(P) = 0$.
\end{definition}

It was proved by Glebskii et al. in \cite{Glebskii:1969} and independently by
Fagin in \cite{Fagin:1976} that the first-order logic has a $0$-$1$ law.
In \cite{Dawar:2010}, Dawar and Gr\"adel extended this result by showing that
the logic $\mathcal{L}^{\omega}_{\infty \omega}$ (the finite variable logic)
augmented with finitely many embedding-closed quantifiers of finite width has a
$0$-$1$ law as well.
We start our investigation of embedding-closed quantifiers by exhibiting this
proof here.
Even though this result is not new, we find it suitable to include it here for
the sake of completeness of our exposition on embedding-closed quantifiers.

A structure $\mathfrak{A}$ is \emph{homogeneous} if every isomorphism between
finitely generated substructures of $\mathfrak{A}$ can be extended to an
automorphism of $\mathfrak{A}$.
Let $\tau$ be a relational vocabulary.
It can be shown that there exists a unique countable homogeneous $\tau$-structure
into which any finite $\tau$-structure can be embedded. We call this structure
the \emph{random $\tau$-structure}.
The following is a well-known fact:

\begin{theorem}
If $P$ is an $\mathcal{L}_{\omega\omega}$-definable property that is true in the
random structure, then $\mu(P) = 1$.
\end{theorem}

\begin{lemma}\label{random}
Let $\tau$ be a finite relational vocabulary,
$\mathfrak{A}$ a finite $\tau$-structure and $\overline{a} \in A^n$ for some
$n<\omega$.
Suppose that $t$ is the atomic type of $\overline{a}$ and
set
\[
	P = \{\mathfrak{B} \in \Str[\tau] \colon
	\text{for all } \overline{b} \in B^n,
	\text{ if } \mathfrak{B} \vDash t(\overline{b})
	\text{ then } (\mathfrak{A},\overline{a}) \leq (\mathfrak{B},\overline{b}) \}.
\]
Then $\mu(P) = 1$.
\end{lemma}
\begin{proof}
Denote the random $\tau$-structure by $\mathfrak{R}$ .
The structure $\mathfrak{A}$ is embeddable into $\mathfrak{R}$ by, say, an
embedding $f$. Let $\overline{b}$ be a tuple in $R$ whose atomic type is $t$.
Since $\mathfrak{R}$ is homogeneous, there is an automorphism $h$ of
$\mathfrak{R}$ that takes the image of $\overline{a}$, denote it by
$f(\overline{a})$, to $\overline{b}$.
Thus, $h \circ f$ is an embedding of $(\mathfrak{A},\overline{a})$ into
$(\mathfrak{R},\overline{b})$, so $P$ holds in the random structure, and since
$P$ is $\mathcal{L}_{\omega\omega}$-definable, we have $\mu(P) = 1$.
\end{proof}

\begin{lemma}[\cite{Dawar:2010}]\label{asympt2}
Let $\tau$ be a finite relational vocabulary, $Q$ embedding-closed quantifier of
finite width $k$ and $\psi_0,\dots,\psi_{k-1}$ quantifier-free $\tau$-formulas.
There is a quantifier-free $\tau$-formula $\vartheta$ such that
$\forall \overline{x}(\vartheta \leftrightarrow Q(\overline{y}_i\psi_i)_{i<k})$
has asymptotic probability $1$.
\end{lemma}

\begin{proof}
Write $\varphi := Q(\overline{y}_i\psi_i)_{i < k}$, and set
\begin{align*}
	\vartheta := \bigvee\{t \colon &t \text{ is an atomic type and }
	(\mathfrak{A},\overline{a}) \vDash t \land \varphi \\
	&\text{ for some finite } \tau \text{-structure } \mathfrak{A}
	\text{ and tuple }\overline{a}\}.
\end{align*}

We clearly have
$\mathfrak{A} \vDash \forall \overline{x}(\varphi \rightarrow \vartheta)$ for
all finite $\tau$-structures $\mathfrak{A}$.
For the other direction, if $\vartheta$ is an empty disjunction, then $\varphi$
defines the empty relation on all finite structures thus being equivalent to a
quantifier-free formula. Therefore, assume that $\vartheta$ is not empty.

For each type $t$ in $\vartheta$, choose a pair
$(\mathfrak{A}_t,\overline{a}_t)$ such that
$(\mathfrak{A}_t,\overline{a}_t) \vDash t \land \varphi$ and $\mathfrak{A}$ is
finite.
Such pairs exist by the definition of $\vartheta$.
Let $\mathfrak{B}$ be a finite $\tau$-structure such that
$\mathfrak{B} \vDash \exists \overline{x}(\vartheta \land \neg \varphi)$.
Then there is a tuple $\overline{b}$ in $B$ such that
$(\mathfrak{B},\overline{b}) \vDash \vartheta \land \neg \varphi$.
Let $t$ be the atomic type of $\overline{b}$.
Then $(\mathfrak{A}_t,\overline{a}_t) \vDash t \land \varphi$, so if
$(\mathfrak{A}_t,\overline{a}_t) \leq (\mathfrak{B},\overline{b})$, then by
Lemma \ref{preservation}, $(\mathfrak{B},\overline{b}) \vDash \varphi$ which is
contradiction.
Thus, $(\mathfrak{A}_t,\overline{a}_t) \nleq (\mathfrak{B},\overline{b})$, so by
Lemma \ref{random}, $\mu(\exists \overline{x}(\vartheta \land \neg \varphi)) = 0$,
so $\mu(\forall \overline{x}(\vartheta \rightarrow \varphi)) = 1$.
\end{proof}

>>>>

\artstatus{am} %%% leave this alone!!  That means you, too!!

\startlocaldefs
\DeclareMathOperator{\emb}{emb}
\endlocaldefs

@@
